---
output:
  word_document: 
    fig_caption: yes
  html_document: default
---



```{r}
getwd()

EUR_USD_dataset = read.csv("Forex Dataset.csv") # historical EUR/USD dataset

```
```{r}
library(dplyr) # A grammar library for data manipulation

EUR_USD_dataset = EUR_USD_dataset %>% 
  rename(
    TICKVOLUME = TICKVOL,
    VOLUME = VOL
    )

View(EUR_USD_dataset)
```
* 1.01. Adding Simple Moving Average(SMA) Trend indicator
```{r}
library(quantmod)#A Library For Quantitative Financial Modelling
library(TTR)
#Simple moving average (SMA)
sma <-SMA(Op(EUR_USD_dataset),from='2018-03-01',to='2022-03-01')
EUR_USD_dataset$smaOPEN <- SMA(Op(EUR_USD_dataset),from='2018-03-01',to='2022-03-01')#for OPENING Price
sma <-SMA(Cl(EUR_USD_dataset),from='2018-03-01',to='2022-03-01')
EUR_USD_dataset$smaCLOSE <- SMA(Cl(EUR_USD_dataset),from='2018-03-01',to='2022-03-01')#for CLOSING Price
sma <-SMA(Hi(EUR_USD_dataset),from='2018-03-01',to='2022-03-01')
EUR_USD_dataset$smaHIGH <- SMA(Hi(EUR_USD_dataset),from='2018-03-01',to='2022-03-01')#for HIGH Price
sma <-SMA(Lo(EUR_USD_dataset),from='2018-03-01',to='2022-03-01')
EUR_USD_dataset$smaLOW <- SMA(Lo(EUR_USD_dataset),from='2018-03-01',to='2022-03-01')#for LOW Price
View(EUR_USD_dataset)
```
* 1.01. Check for missing values (NA's) in the records using the summary function.
* 1.02. Use Data Explorer packages to further analyze records.
```{r}
summary(EUR_USD_dataset)
library(DataExplorer) # A library to automate visual exploration of data 
plot_intro(EUR_USD_dataset)
plot_missing(EUR_USD_dataset)

library(mice) # A library for Multivariate Imputation by Chained Equations
md.pattern(EUR_USD_dataset)
mice_imputes = mice(EUR_USD_dataset, m = 20, maxit = 40, seed = 10) 
mice_imputes$method

EUR_USD_dataset = complete(mice_imputes, 20) 

md.pattern(EUR_USD_dataset)
summary(EUR_USD_dataset) #confirm absence of missing data and gain insights on the data
View(EUR_USD_dataset)
#Missing Data Present
#since number is negligible, we effect complete case analysis
CleanSet = na.omit(EUR_USD_dataset)
summary(CleanSet)#confirm absence of missing data and gain insights on the data
plot_missing(CleanSet)
View(CleanSet)
str(CleanSet)
CleanSet$TICKVOLUME = as.numeric(as.integer(CleanSet$TICKVOLUME))
CleanSet$VOLUME = as.numeric(as.integer(CleanSet$VOLUME))
CleanSet$SPREAD = as.numeric(as.integer(CleanSet$SPREAD))
CleanSet$BULL_BEAR = as.numeric(as.integer(CleanSet$BULL_BEAR))
str(CleanSet)
```

#### Univariate Analyses for market Features

*Evaluating standard deviation and kurtosis for each feature
```{r message=TRUE, warning=TRUE, eval=FALSE}
library(moments)#A library to visualize our  features

#Histogram for Open
print(paste0("standard deviation: ", sd(CleanSet$OPEN)))
print(paste0("Kurtosis: ", round(kurtosis(CleanSet$OPEN),2)))

options(repr.plot.width = 4, repr.plot.height = 4)
hist(CleanSet$OPEN, breaks=100, prob=TRUE) # Make it a probability distribution

m<-mean(CleanSet$OPEN)
std<-sqrt


#Histogram for High
print(paste0("standard deviation: ", sd(CleanSet$HIGH)))
print(paste0("Kurtosis: ", round(kurtosis(CleanSet$HIGH),2)))

options(repr.plot.width = 4, repr.plot.height = 4)
hist(CleanSet$HIGH, breaks=100, prob=TRUE) # Make it a probability distribution

m<-mean(CleanSet$HIGH)
std<-sqrt(var(CleanSet$HIGH))


#Histogram for Low
print(paste0("standard deviation: ", sd(CleanSet$LOW)))
print(paste0("Kurtosis: ", round(kurtosis(CleanSet$LOW),2)))

options(repr.plot.width = 4, repr.plot.height = 4)
hist(CleanSet$LOW, breaks=100, prob=TRUE) # Make it a probability distribution

m<-mean(CleanSet$LOW)
std<-sqrt(var(CleanSet$LOW))


#Histogram for Close
print(paste0("standard deviation: ", sd(CleanSet$CLOSE)))
print(paste0("Kurtosis: ", round(kurtosis(CleanSet$CLOSE),2)))

options(repr.plot.width = 4, repr.plot.height = 4)
hist(CleanSet$CLOSE, breaks=100, prob=TRUE) # Make it a probability distribution

m<-mean(CleanSet$CLOSE)
std<-sqrt(var(CleanSet$CLOSE))


#Histogram for TickVolume
print(paste0("standard deviation: ", sd(CleanSet$TICKVOLUME)))
print(paste0("Kurtosis: ", round(kurtosis(CleanSet$TICKVOLUME),2)))

options(repr.plot.width = 4, repr.plot.height = 4)
hist(CleanSet$TICKVOLUME, breaks=100, prob=TRUE) # Make it a probability distribution
m<-mean(CleanSet$TICKVOLUME)
std<-sqrt(var(CleanSet$TICKVOLUME))


#Histogram for Volume
print(paste0("standard deviation: ", sd(CleanSet$VOLUME)))
print(paste0("Kurtosis: ", round(kurtosis(CleanSet$VOLUME),2)))

options(repr.plot.width = 4, repr.plot.height = 4)
hist(CleanSet$VOLUME, breaks=100, prob=TRUE) # Make it a probability distribution

m<-mean(CleanSet$VOLUME)
std<-sqrt(var(CleanSet$VOLUME))


#Histogram for Spread
print(paste0("standard deviation: ", sd(CleanSet$SPREAD)))
print(paste0("Kurtosis: ", round(kurtosis(CleanSet$SPREAD),2)))

options(repr.plot.width = 4, repr.plot.height = 4)
hist(CleanSet$SPREAD, breaks=100, prob=TRUE) # Make it a probability distribution

m<-mean(CleanSet$SPREAD)
std<-sqrt(var(CleanSet$SPREAD))


#Histogram for Bull_Bear
print(paste0("standard deviation: ", sd(CleanSet$BULL_BEAR)))
print(paste0("Kurtosis: ", round(kurtosis(CleanSet$BULL_BEAR),2)))

options(repr.plot.width = 4, repr.plot.height = 4)
hist(CleanSet$BULL_BEAR, breaks=100, prob=TRUE) # Make it a probability distribution

m<-mean(CleanSet$BULL_BEAR)
std<-sqrt(var(CleanSet$BULL_BEAR))


#Histogram for smaOpen
print(paste0("standard deviation: ", sd(CleanSet$smaOPEN)))
print(paste0("Kurtosis: ", round(kurtosis(CleanSet$smaOPEN),2)))

options(repr.plot.width = 4, repr.plot.height = 4)
hist(CleanSet$smaOPEN, breaks=100, prob=TRUE) # Make it a probability distribution

m<-mean(CleanSet$smaOPEN)
std<-sqrt(var(CleanSet$smaOPEN))


#Histogram for smaHigh
print(paste0("standard deviation: ", sd(CleanSet$smaHIGH)))
print(paste0("Kurtosis: ", round(kurtosis(CleanSet$smaHigh),2)))

options(repr.plot.width = 4, repr.plot.height = 4)
hist(CleanSet$smaHIGH, breaks=100, prob=TRUE) # Make it a probability distribution

m<-mean(CleanSet$smaHigh)



#Histogram for smaLow
print(paste0("standard deviation: ", sd(CleanSet$smaLOW)))
print(paste0("Kurtosis: ", round(kurtosis(CleanSet$smaLOW),2)))

options(repr.plot.width = 4, repr.plot.height = 4)
hist(CleanSet$smaLOW, breaks=100, prob=TRUE) # Make it a probability distribution

m<-mean(CleanSet$smaLOW)
std<-sqrt(var(CleanSet$smaLOW))

#Histogram for smaClose

print(paste0("standard deviation: ", sd(CleanSet$smaCLOSE)))
print(paste0("Kurtosis: ", round(kurtosis(CleanSet$smaCLOSE),2)))

options(repr.plot.width = 4, repr.plot.height = 4)
hist(CleanSet$smaCLOSE, breaks=100, prob=TRUE) # Make it a probability distribution

m<-mean(CleanSet$smaCLOSE)
std<-sqrt(var(CleanSet$smaCLOSE))
# Overlay a standard normal distribution
curve(dnorm(x, mean=m, sd=std), col="darkblue", lwd=2, add=TRUE )
```

* Median opening price is within opening price 1.10 and 1.15 

* The kurtosis measure describes the tail of a distribution, how similar are the outlying values of the distribution to the standard normal distribution.
The standard normal distribution has a kurtosis of 0.
A negative value for kurtosis indicates a thin tailed distribution; the values of the sample are distributed closer to the median than we would expect for a standard normal distribution. A positive kurtosis value indicates we are dealing with a fat tailed distribution, where extreme outcomes are more common than would be predicted by a standard normal distribution.
Fat-tailed distribution are particularly interesting as it indicates that the market has the capacity to generate significant extreme values that don’t fall into the standard normal distribution.
From above analysis we can see the price movements largely is fat-tailed. 
*Our Kurtosis Test indicate strong presence of outliers





#### Bivariate Analysis on the dataset

* 1.2. We go on to chart a scatterplot to see the effect of independent variables on the probability of BULL_BEAR

```{r}
library(ggplot2)#A library for plotting our scatterplot
#OPEN
plot(CleanSet$OPEN, CleanSet$BULL_BEAR, pch=16, col='steelblue',
     main='BULL_BEAR vs. OPEN',
     xlab='BULL_BEAR', ylab='OPEN')
#smaOPEN
plot(CleanSet$smaOPEN, CleanSet$BULL_BEAR, pch=16, col='steelblue',
     main='BULL_BEAR vs. smaOPEN',
     xlab='BULL_BEAR', ylab='smaOPEN')
#CLOSE
plot(CleanSet$CLOSE, CleanSet$BULL_BEAR, pch=16, col='steelblue',
     main='BULL_BEAR vs. CLOSE',
     xlab='BULL_BEAR', ylab='CLOSE')
#smaCLOSE
plot(CleanSet$smaCLOSE, CleanSet$BULL_BEAR, pch=16, col='steelblue',
     main='BULL_BEAR vs. smaCLOSE',
     xlab='BULL_BEAR', ylab='smaCLOSE')
#HIGH
plot(CleanSet$HIGH, CleanSet$BULL_BEAR, pch=16, col='steelblue',
     main='BULL_BEAR vs. HIGH',
     xlab='BULL_BEAR', ylab='HIGH')
#smaHIGH
plot(CleanSet$smaHIGH, CleanSet$BULL_BEAR, pch=16, col='steelblue',
     main='BULL_BEAR vs. smaHIGH',
     xlab='BULL_BEAR', ylab='smaHIGH')
#TICKVOLUME
plot(CleanSet$TICKVOLUME, CleanSet$BULL_BEAR, pch=16, col='steelblue',
     main='BULL_BEAR vs. TICKVOLUME',
     xlab='BULL_BEAR', ylab='TICKVOLUME')
#VOLUME
plot(CleanSet$VOLUME, CleanSet$BULL_BEAR, pch=16, col='steelblue',
     main='BULL_BEAR vs. VOLUME',
     xlab='BULL_BEAR', ylab='VOLUME')
#SPREAD
plot(CleanSet$SPREAD, CleanSet$BULL_BEAR, pch=16, col='steelblue',
     main='BULL_BEAR vs. SPREAD',
     xlab='BULL_BEAR', ylab='SPREAD')
#LOW
plot(CleanSet$LOW, CleanSet$BULL_BEAR, pch=16, col='steelblue',
     main='BULL_BEAR vs. LOW',
     xlab='BULL_BEAR', ylab='LOW')
#smaLOW
plot(CleanSet$smaLOW, CleanSet$BULL_BEAR, pch=16, col='steelblue',
     main='BULL_BEAR vs. smaLOW',
     xlab='BULL_BEAR', ylab='smaLOW')


```

* 1.2. No cogent inference was drawn from the bivarete analysis except that the dependent correlation shows a positive correlation with the independent variables

* 1.66. Use the psych package to carry out multivariate analysis
* 1.67. Carry out correlation analysis on the dataset


* By default corr.test produces pairwise "Pearson" correlation matrix for the entire dataset
```{r}
library(psych) # multivariate analysis

NewCleanSet = CleanSet[,-c(1,5,11)]#data set without date,close and smaCLOSE
corr.test(NewCleanSet) 

```
#### Correlation test findings 

* All variables showed very low correlation to BULL_BEAR
  


* 1.69. Check significance of correlations 
```{r}
# Check the correlation significance levels

cor.test(NewCleanSet$BULL_BEAR,NewCleanSet$OPEN)$p.value
cor.test(NewCleanSet$BULL_BEAR,NewCleanSet$HIGH)$p.value 
cor.test(NewCleanSet$BULL_BEAR,NewCleanSet$LOW)$p.value 
cor.test(NewCleanSet$BULL_BEAR,NewCleanSet$TICKVOLUME)$p.value 
cor.test(NewCleanSet$BULL_BEAR,NewCleanSet$VOLUME)$p.value
cor.test(NewCleanSet$BULL_BEAR,NewCleanSet$SPREAD)$p.value
cor.test(NewCleanSet$BULL_BEAR,NewCleanSet$smaOPEN)$p.value
cor.test(NewCleanSet$BULL_BEAR,NewCleanSet$smaHIGH)$p.value
cor.test(NewCleanSet$BULL_BEAR,NewCleanSet$smaLOW)$p.value

```
* There exists correlation significance between BULL_BEAR and each of all the variables except VOLUME and TICKVOLUME
  
  
*
### 2.0. Machine Learning Analysis 

* 2.1. Load mlbench and caret packages respectively
* 2.2. Ensure that dependent variable "BULL_BEAR" is of the factor class

```{r}
library(mlbench)
library(caret)
str(NewCleanSet)
sapply(NewCleanSet,class)
NewCleanSet$BULL_BEAR = as.factor(as.integer(NewCleanSet$BULL_BEAR))
summary(NewCleanSet)

```

* 2.3. Perform feature selection. Correlation analyses findings in Section 1 already pointed to the fact that some
       features have no association with the outcome of the dependent variable.
* 2.4. Feature Selection using Boruta, mlbench and Caret packages respectively

```{r}
library(Boruta)

# Feature Selection operation
set.seed(7)
boruta = Boruta(NewCleanSet$BULL_BEAR~.,data = NewCleanSet, doTrace = 4, maxRuns = 100)
plot(boruta, las = 2, cex.axis = 0.5)

# reclassify features that were designated as tentative
bor = TentativeRoughFix(boruta)
plot(bor, las = 2, cex.axis = 0.5)

attStats(boruta)
print(bor)

# create object and limit features to only the 5 confirmed features from the boruta analysis + the dependent variable

col_order2 = c("OPEN","HIGH","LOW","smaOPEN","smaLOW","BULL_BEAR")

NewSet = NewCleanSet[, col_order2]
View(NewSet)
str(NewSet)
levels(NewSet)
```
* The feature selection operation using the Boruta package works as follows:

* Firstly, it introduces randomness into the data set by producing shuffled copies of all the features of 
  interest (these are referred to as shadow features).

* Next, it trains a random forest classifier on the extended data set and applies a feature importance measure 
  (the default is Mean Decrease Accuracy) to ascertain the relevance of each of the features. Those with higher 
  means are classified as more important.

* Afterwards, at every given iteration, it checks whether a real feature has a higher importance than the best 
  of its shadow features (in other words, it checks to see if the feature has a higher Z-score than the maximum 
  Z-score of its shadow features) and therefore eliminates features which are termed highly irrelevant.

*  Lastly, the Boruta algorithm stops either when all features gets confirmed or rejected or when it reaches a 
  specified limit of random forest runs
  
  * 2.5. Split the NewSet features object into training and validation groups (objects)
* 2.6. Split out validation dataset
* 2.7. Do this by creating a list of 70% of the rows in the original dataset to serve as training set

```{r}
#LOW/BEARISH TREND
LOW = NewSet[NewSet$BULL_BEAR == 0,]

#HIGH/BULLISH TREND
HIGH = NewSet[NewSet$BULL_BEAR == 1,]

set.seed(7)
validationIndex = createDataPartition(NewSet$BULL_BEAR , p=0.70, list = FALSE)

#select 30% of the data for validation

validation = NewSet[-validationIndex,]

#use the remaining 70% of data to train the model
dataset = NewSet[validationIndex,]

# Check that the distribution of the dependent variable is similar in train and test sets
prop.table(table(NewSet$BULL_BEAR))
prop.table(table(dataset$BULL_BEAR))
prop.table(table(validation$BULL_BEAR))
```
* The derived figures show that the ratio of the "LOW":"HIGH" was proportionally split

* 2.8. Evaluate 8 Machine Learning Algorithms using  our Historical forex data
* 2.9. Linear Algorithms - Generalized linear model (LG) and Linear Discriminate Analysis (LDA)
* 2.9.1 Non-Linear Algorithms - K-Nearest Neighbors (KNN),Support Vector Machine(SVM), 
        Neural Network(pcaNNEt),Multiple Layer Perceptron(MLP)
* 2.9.2 ENSEMBLE METHODS: Random Forest (RF), Naive Bayes,
* 2.9.3 Analysis is done using 10-fold cross validation with 3 repeats

```{r}
trainControl = trainControl(method = "repeatedcv", number = 10, repeats = 3)
metric = "Accuracy"

#1 LG
set.seed(7)
fit.glm = train(BULL_BEAR~., data=dataset, method="glm", metric=metric, trControl=trainControl)

#2 LDA
set.seed(7)
fit.lda = train(BULL_BEAR~., data=dataset, method="lda", metric=metric, trControl=trainControl)

#3 KNN
set.seed(7)
fit.knn = train(BULL_BEAR~., data=dataset, method="knn", metric=metric, trControl=trainControl)


#4 Random Forest
set.seed(7)
fit.rf = train(BULL_BEAR~., data=dataset, method="rf", metric=metric, trControl=trainControl)

#5 SVM
set.seed(7)
fit.svmLinear = train(BULL_BEAR~., data=dataset, method="svmLinear", metric=metric, trControl=trainControl)

#6 Naive Bayes
set.seed(7)
fit.naive_bayes = train(BULL_BEAR~., data=dataset, method="naive_bayes", metric=metric, trControl=trainControl)

#7 Multi Layer Perceptron
set.seed(7)
fit.mlp = train(BULL_BEAR~., data=dataset, method="mlp", metric=metric, trControl=trainControl)

#8 Neural Network
set.seed(7)
fit.pcaNNet = train(BULL_BEAR~., data=dataset, method="pcaNNet", metric=metric, trControl=trainControl)


# Compare algorithms
results = resamples(list(LG=fit.glm, LDA=fit.lda, KNN=fit.knn,RF=fit.rf,  SVM=fit.svmLinear, NB = fit.naive_bayes, MLP = fit.mlp, NN=fit.pcaNNet)) 
                          
summary(results)
dotplot(results)
```

####* Top three performing algorithms are LG,LDA SVM, with mean accuracies of 0.8628, 0.8619 and 0.8416 respectively


#### Tune the top two learning algorithms

* 4.18. Tuning for the LG learning algorithm

```{r}
trainControl = trainControl(method="repeatedcv", number=10, repeats=3) 
metric = "Accuracy"

#LG
set.seed(7)
grid = expand.grid(.parameter = seq(1,10, by=1))
fit.LG_tuned = train(BULL_BEAR~.,data = dataset, method="glm", metric=metric, tuneGrid=grid, trControl=trainControl)
fit.LG_tuned
```
* Tuning did not improve the LG model


* 4.19. Tuning for the LDA learning algorithm

```{r}
trainControl = trainControl(method="repeatedcv", number=10, repeats=3) 
metric = "Accuracy"

#LDA
set.seed(7)
grid = expand.grid(.parameter = seq(1,10, by=1))
fit.lda_tuned = train(BULL_BEAR~.,data = dataset, method="lda", metric=metric, tuneGrid=grid, trControl=trainControl)
fit.lda_tuned
```
* Tuning did not improve the LDA model



* 3. Finalize by building Model using LG
* 3.1. Prepare the data transform for finalizing Model 

```{r}
set.seed(7)
x = dataset[,1:6]
preprocessParam = preProcess(x)
x = predict(preprocessParam, x)

# prepare the validation dataset
set.seed(7)

# transform the validation dataset
validationX = predict(preprocessParam, validation[,1:6])
```
* 3.2. Make predictions and build confusion matrix for Model 1


```{r}
glmPredict = predict(fit.glm, newdata = validation)
confusionMatrix(glmPredict, validation$BULL_BEAR)
```

* Model performed at an accuracy of 0.8697 with a specificity of 0.8742 with a 
  sensitivity score of 0.8654. 

* 3.3. Calculate AUC(ROC) for Model 1

```{r}
library(PRROC)
sapply(glmPredict, class)
sapply(validation, class)
pred = as.numeric(as.character(glmPredict))
valid = as.character(as.factor(validation$BULL_BEAR))
valid = as.numeric(as.character(validation$BULL_BEAR))
prroc_obj = roc.curve(scores.class0 = pred, weights.class0 = valid, curve = TRUE)
plot(prroc_obj)
```


 